---
title: "Survey analysis"
output:
  html_document:
    code_folding: hide
    fig_width: 8
    fig_height: 4
    toc: true
    toc_depth: 3
    toc_float: true
theme: sandstone
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(tidyverse)
library(lmerTest)
library(effects)
library(patchwork)
library(psycho)
library(report)
# setwd("C:/Users/mwillems/OneDrive/papers/explRecSys/data")


# If using brewer: scale_fill_manual(values = plot_palette)
# If manual: scale_fill_manual(values = plot_palette)
plot_palette <- c("#2596be", "#93cbac") #"Paired"
plot_base_size <- 16
plot_bar_width <- 0.6
plot_errorbar_dodge <- 0.6
plot_errorbar_width <- 0.3
show.report <- \(x) print(report(x, CI = 95))
```

```{r reading, include=FALSE}
participants <- read_csv("participants.csv")
evaluations <- read_csv("evaluations.csv")
recommendations <- read_csv("recommendations.csv")
preferences <- read_csv("preferences.csv")
```

```{r filtering, include=FALSE}
participants <- participants %>% filter(!startsWith(prolificPID, "NON-PROLIFIC"))
evaluations <- evaluations %>% filter(!startsWith(prolificPID, "NON-PROLIFIC"))
recommendations <- recommendations %>% filter(!startsWith(prolificPID, "NON-PROLIFIC"))
preferences <- preferences %>% filter(!startsWith(prolificPID, "NON-PROLIFIC"))
```

```{r, include=FALSE}
completeids <- participants %>% filter(status == "finished") %>% pull(prolificPID)
```

```{r, include=FALSE}
participants <- participants %>% filter(prolificPID %in% completeids)
evaluations <- evaluations %>% filter(prolificPID %in% completeids)
recommendations <- recommendations %>% filter(prolificPID %in% completeids)
preferences <- preferences %>% filter(prolificPID %in% completeids)
```

```{r}
evaluations_detailed <- evaluations %>%
  left_join(
    recommendations %>%
      select(prolificPID, recID, shouldWatch, userBasedExp, recommender),
    by = c("prolificPID", "recID")
  ) %>%
  pivot_wider(
    id_cols = c(prolificPID, recID, shouldWatch, userBasedExp, recommender),
    names_from = questionNumber,
    values_from = response,
    names_prefix = "Q"
  ) %>%
  rename(know = Q1) %>%
  mutate(know = ifelse(know=="true", "Known", "Unknown"),
         recType = recommender,
         expType = ifelse(userBasedExp, "User-based", "Generic"),
         across(starts_with("Q"), as.numeric))
```

```{r}
dfnl <- evaluations_detailed %>% filter(!shouldWatch)
df <- evaluations_detailed %>% filter(shouldWatch)
```

# RQ1 - How do users evaluate ChatGPT-generated recommendations compared to random (but popular) recommendations?

## For recommendations

```{r}
M2<-lmer(Q2~recType*know+(1|prolificPID),data=df)
summary(M2)
```

```{r, results='asis'}
show.report(M2)
```

```{r}
ef <- effect("recType:know", M2)
m2i<-as.data.frame(ef)
p1 <- ggplot(m2i, aes(x = recType, y = fit, fill = know)) +
  geom_bar(stat = "identity", position = "dodge", width = plot_bar_width) +
  geom_errorbar(aes(ymin = fit - se, ymax = fit + se),
                width = plot_errorbar_width,
                position = position_dodge(width = 0.6)) +
  theme_bw(base_size = plot_base_size) +
  labs(y = "Satisfaction",
       title = "(a) Question: I would enjoy this recommendation.",
       fill = "Previous knowledge about the movie",
       x = "Recommender") +coord_cartesian(ylim=c(1,5))+
  theme(legend.position = "top") +
  scale_fill_manual(values = plot_palette)

p1
```

## For disrecommendations (do not report this in the paper yet)

```{r}
M2d<-lmer(Q2~know+(1|prolificPID),data=dfnl)
summary(M2d)
```

```{r, results='asis'}
show.report(M2d)
```

# RQ2 - How do users perceive personalized versus generic ChatGPT-generated explanations regarding the recommendation method, effectiveness, and persuasiveness of explanations for recommendations?

## For recommendations

### Helpfulness

```{r}
M3<-lmer(Q3~recType*expType+(1|prolificPID),data=df)
summary(M3)
```

```{r, results='asis'}
show.report(M3)
```

```{r}
ef <- effect("recType:expType", M3)
m3i<-as.data.frame(ef)
p2 <- ggplot(m3i, aes(recType, fit, fill = expType)) +
  geom_bar(stat = "identity", position = "dodge", width = plot_bar_width) + 
  geom_errorbar(aes(ymin = fit - se, ymax = fit + se),
                width = plot_errorbar_width,
                position = position_dodge(width = plot_errorbar_dodge)) +
  theme_bw(base_size = plot_base_size) + 
  labs(y = "Effectiveness",
       title = "(b) Question: This explanation helps me to \ndetermine how well I will like this movie.",
       x = "Recommender",
       fill = "Explanation Type") +
  theme(legend.position = "top") +coord_cartesian(ylim=c(1,5))+
  scale_fill_manual(values = plot_palette)

p2
```

### Personalization

```{r}
M4<-lmer(Q4~recType*expType+(1|prolificPID),data=df)
summary(M4)
```

```{r, results='asis'}
show.report(M4)
```


```{r}
ef <- effect("recType:expType", M4)
m4i<-as.data.frame(ef)
p3 <- ggplot(m4i, aes(recType, fit, fill = expType)) +
  geom_bar(stat = "identity", position = "dodge", width = plot_bar_width) +
  geom_errorbar(aes(ymin = fit - se, ymax = fit + se),
                width = plot_errorbar_width,
                position = position_dodge(width = plot_errorbar_dodge)) +
  theme_bw(base_size = plot_base_size) +
  labs(y = "Personalization",
       title = "(c) Question: This explanation resonates well with\naspects of movies that I like.",
       x = "Recommender",
       fill = "Explanation Type") +
  theme(legend.position = "top") +coord_cartesian(ylim=c(1,5))+
  scale_fill_manual(values = plot_palette)

p3
```

### Persuasiveness

```{r}
M5<-lmer(Q5~recType*expType+(1|prolificPID),data=df)
summary(M5)
```

```{r, results='asis'}
show.report(M5)
```

```{r}
ef <- effect("recType:expType", M5)
m5ai<-as.data.frame(ef)
p4 <- ggplot(m5ai, aes(recType, fit, fill = expType)) +
  geom_bar(stat = "identity", position = "dodge", width = plot_bar_width) +
  geom_errorbar(aes(ymin = fit - se, ymax = fit + se),
                width = plot_errorbar_width,
                position = position_dodge(width = plot_errorbar_dodge)) +
  theme_bw(base_size = plot_base_size) +
  labs(title = "(d) Question: This explanation is convincing.",
       y = "Persuasiveness",
       x = "Recommender",
       fill = "Explanation Type") +
  theme(legend.position = "top") +coord_cartesian(ylim=c(1,5))+
  scale_fill_manual(values = plot_palette)

p4
```

```{r, fig.height=9, fig.width=16}
(p1 + theme(plot.margin = unit(c(0,30,30,0), "pt")) + p2) / (p3 + p4)
```

```{r}
ggsave("plot-recs.pdf", device = "pdf", width = 16, height = 9)
```

## For disrecommendations

### Helpfulness

```{r}
M3d<-lmer(Q3~expType+(1|prolificPID),data=dfnl)
summary(M3d)
```

```{r, results='asis'}
show.report(M3d)
```

### Personalization

```{r}
M4d<-lmer(Q4~expType+(1|prolificPID),data=dfnl)
summary(M4d)
```

```{r, results='asis'}
show.report(M4d)
```

### Persuasiveness

```{r}
M5d<-lmer(Q5~expType*know+(1|prolificPID),data=dfnl)
summary(M5d)
```

```{r, results='asis'}
show.report(M5d)
```

# RQ4 - Do personalized versus generic explanations work differently for familiar or unfamiliar movies?

## For recommendations

### Helpfulness

```{r}
M3<-lmer(Q3~recType+expType+ expType*know + (1|prolificPID),data=df)
summary(M3)
```

```{r, results='asis'}
show.report(M3)
```

```{r}
ef <- effect("recType:expType:know", M3)
m3i<-as.data.frame(ef)
p2 <- ggplot(m3i, aes(recType, fit, fill = expType)) +
  geom_bar(stat = "identity", position = "dodge", width = plot_bar_width) +
  geom_errorbar(aes(ymin = fit - se, ymax = fit + se),
                width = plot_errorbar_width,
                position = position_dodge(width = plot_errorbar_dodge)) +
  theme_bw(base_size = plot_base_size) +
  facet_wrap(~ know) +
  labs(y = "Effectiveness",
       title = "(a) Question: This explanation helps me\nto determine how well I will like this\nmovie.",
       x = "Recommender",
       fill = "Explanation Type") +
  theme(legend.position = "top") +coord_cartesian(ylim=c(1,5))+
  scale_fill_manual(values = plot_palette)

p2
```

### Personalization

```{r}
M4<-lmer(Q4~recType*expType+ expType*know+(1|prolificPID),data=df)
summary(M4)
```

```{r, results='asis'}
show.report(M4)
```


```{r}
ef <- effect("recType:expType:know", M4)
m4i<-as.data.frame(ef)

m4i

p3 <- ggplot(m4i, aes(recType, fit, fill = expType)) +
  geom_bar(stat = "identity", position = "dodge", width = plot_bar_width) +
  geom_errorbar(aes(ymin = fit - se, ymax = fit + se),
                width = plot_errorbar_width,
                position = position_dodge(width = plot_errorbar_dodge)) +
  theme_bw(base_size = plot_base_size) +
  facet_wrap(~ know) +
  labs(y = "Personalization",
       title = "(b) Question: This explanation resonates well\nwith aspects of movies that I like.",
       x = "Recommender",
       fill = "Explanation Type") +
  theme(legend.position = "top") +coord_cartesian(ylim=c(1,5))+
  scale_fill_manual(values = plot_palette)

p3
```

### Persuasiveness

```{r}

#M5<-lmer(Q5~recType*expType*know+(1|prolificPID),data=df)
M5<-lmer(Q5~recType*expType + expType*know+(1|prolificPID),data=df)
summary(M5)
```

```{r, results='asis'}
show.report(M5)
```

```{r}
ef <- effect("recType:expType:know", M5)
m5ai<-as.data.frame(ef)
p4 <- ggplot(m5ai, aes(recType, fit, fill = expType)) +
  geom_bar(stat = "identity", position = "dodge", width = plot_bar_width) +
  geom_errorbar(aes(ymin = fit - se, ymax = fit + se),
                width = plot_errorbar_width,
                position = position_dodge(width = plot_errorbar_dodge)) +
  theme_bw(base_size = plot_base_size) + coord_cartesian(ylim=c(1,5))+
  facet_wrap(~ know) +
  labs(title = "(c) Question: This explanation is\nconvincing.",
       y = "Persuasiveness",
       x = "Recommender",
       fill = "Explanation Type") +
  theme(legend.position = "top") +
  scale_fill_manual(values = plot_palette)

p4
```

```{r, fig.height=5, fig.width=16}
(p2 + theme(plot.margin = unit(c(0,10,0,0), "pt"))) + (p3 + theme(plot.margin = unit(c(0,10,0,0), "pt"))) + p4 +
  plot_layout(guides = 'collect') &
  theme(legend.position = "top")
```

```{r}
ggsave("plot-familiarity.pdf", device = "pdf", width = 16, height = 5)
```

## For disrecommendations

### Helpfulness

```{r}
M3d<-lmer(Q3~expType*know+(1|prolificPID),data=dfnl)
summary(M3d)
```

```{r, results='asis'}
show.report(M3d)
```

```{r}
ef <- effect("expType:know", M3d)
m3i<-as.data.frame(ef)

p5 <- ggplot(m3i, aes(know, fit, fill = expType)) +
  geom_bar(stat = "identity", position = "dodge", width = plot_bar_width) +
  geom_errorbar(aes(ymin = fit - se, ymax = fit + se),
                width = plot_errorbar_width,
                position = position_dodge(width = plot_errorbar_dodge)) +
  theme_bw(base_size = plot_base_size) +
  labs(y = "Effectiveness",
       title = "(a) Question: This explanation helps me\nto determine how well I will dislike this\nmovie.",
       x = "Previous knowledge about the movie",
       fill = "Explanation Type") +
  theme(legend.position = "top") +coord_cartesian(ylim=c(1,5))+
  scale_fill_manual(values = plot_palette)

p5
```
user-based significantly lower if we clean up the model
```{r}
M3d2<-lmer(Q3~expType+(1|prolificPID),data=dfnl)
summary(M3d2)
```


### Personalization

```{r}
M4d<-lmer(Q4~expType*know+(1|prolificPID),data=dfnl)
summary(M4d)
```

```{r, results='asis'}
show.report(M4d)
```

```{r}
ef <- effect("expType:know", M3d)
m4i<-as.data.frame(ef)

p6 <- ggplot(m4i, aes(know, fit, fill = expType)) +
  geom_bar(stat = "identity", position = "dodge", width = plot_bar_width) +
  geom_errorbar(aes(ymin = fit - se, ymax = fit + se),
                width = plot_errorbar_width,
                position = position_dodge(width = plot_errorbar_dodge)) +
  theme_bw(base_size = plot_base_size) +
  labs(y = "Personalization",
       title = "(b) Question: This explanation resonates well\nwith aspects of movies that I dislike.",
       x = "Previous knowledge about the movie",
       fill = "Explanation Type") +
  theme(legend.position = "top") +coord_cartesian(ylim=c(1,5))+
  scale_fill_manual(values = plot_palette)

p6
```
#without interaction of known expType still not significant
```{r}
M4d2<-lmer(Q4~expType+(1|prolificPID),data=dfnl)
summary(M4d2)
```
### Persuasiveness

```{r}
M5d<-lmer(Q5~expType*know+(1|prolificPID),data=dfnl)
summary(M5d)
```

```{r, results='asis'}
show.report(M5d)
```

```{r}
ef <- effect("expType:know", M5d)
m5i <- as.data.frame(ef)
p7 <- ggplot(m5i, aes(x = know, y = fit, fill = expType)) +
  geom_bar(stat = "identity", position = "dodge", width = plot_bar_width) +
  geom_errorbar(aes(ymin = fit - se, ymax = fit + se),
                width = plot_errorbar_width,
                position = position_dodge(width = plot_errorbar_dodge)) +
  theme_bw(base_size = plot_base_size) +
  labs(title = "(c) Question: This explanation is\nconvincing.",
       y = "Persuasiveness",
       x = "Previous knowledge about the movie",
       fill = "Explanation Type") +coord_cartesian(ylim=c(1,5))+
  theme(legend.position = "top") +
  scale_fill_manual(values = plot_palette)

p7
```
```{r}
M5d2<-lmer(Q5~expType+(1|prolificPID),data=dfnl)
summary(M5d2)
```

```{r, fig.height=5, fig.width=16}
(p5 + theme(plot.margin = unit(c(0,10,0,0), "pt"))) + (p6 + theme(plot.margin = unit(c(0,10,0,0), "pt"))) + p7 +
  plot_layout(guides = 'collect') &
  theme(legend.position = "top")
```
```{r}
ggsave("plot-disrecs.pdf", device = "pdf", width = 16, height = 5)
```


```{r}
#convert to csv for mplus
dtm<-df %>% mutate(id=as.numeric(as.factor(prolificPID)), userBased=ifelse(userBasedExp, 1, 0), GPTrec=ifelse(	
recommender=="random-imdb",0,1), enjoy=Q2, eff=Q3,personalized=Q4, persuasive=Q5, known=ifelse(know=="Known", 1,0))
write.table(dtm[,13:20],"explrec.txt", sep=" ", row.names=F)

```

